---
title: "Diagnostic performance of left ventricular mechanical dyssynchrony indices using cardiovascular magnetic resonance feature tracking"
author:
 - "Daniel E Loewenstein MD^1^, Björn Wieslander MD PhD^1^, Einar Heiberg PhD^2^, Jimmy Axelsson MD^1^, Igor Klem MD^3^, Robin Nijveldt MD PhD^4^, Erik B Schelbert MSc MD^5^, Peder Sörensson MD PhD^6^, Andreas Sigfridsson PhD^1^, David G Strauss MD^7^,\nRaymond J Kim MD^8^, Brett D Atwater MD^3,9^, Martin Ugander MD PhD^1,10^"
 - "^1^Department of Clinical Physiology, Karolinska Institutet, and Karolinska University Hospital, Solna, Stockholm, Sweden"
 - "^2^Clinical Physiology, Department of Clinical Sciences Lund, Lund University, Skåne University Hospital, Lund, Sweden"
 - "^3^Division of Cardiology, Duke University Medical Center, Durham, NC, USA"
 - "^4^Department of Cardiology, Radboud University Medical Center, Nijmegen, the Netherlands"
 - "^5^Minneapolis Heart Institute East, Allina Health, United Hospital, St. Paul, MN, USA"
 - "^6^Department of Medicine Solna, Karolinska Institutet, and Department of Cardiology, Karolinska University Hospital, Solna, Stockholm, Sweden"
 - "^7^Division of Applied Regulatory Science, Office of Clinical Pharmacology, Office of Translational Sciences, Center for Drug Evaluation and Research, U.S. Food and Drug Administration, Silver Spring, MD, USA"
 - "^8^Duke Cardiovascular Magnetic Resonance Center, Duke University Medical Center, Durham, NC, USA"
 - "^9^Section of Cardiac Electrophysiology, Inova Heart and Vascular Institute, Fairfax, VA, USA"
 - "^10^Kolling Institute, Royal North Shore Hospital, and University of Sydney, Sydney, Australia"
csl: auxilliary/vancouver.csl
bibliography: auxilliary/library-bibtex-version.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```

```{r}
# Set paths
PROJECT_DIR <- rprojroot::find_rstudio_root_file()
RAW_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "raw_data")
DERIVED_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "derived_data")
FIG_DIR <- file.path(PROJECT_DIR, "analysis", "figures")
```

```{r}
dillacs <- readRDS(file.path(DERIVED_DATA_DIR, "dyssynchrony_data.Rds"))
stats <- readRDS(file.path(DERIVED_DATA_DIR, "statistics.Rds"))
figs  <- readRDS(file.path(DERIVED_DATA_DIR, "figs.Rds"))
source(file.path(PROJECT_DIR, "R", "offset2groups.R"))
```

```{r}
pval_print <- function(x) {
    if (x < 0.001)
        paste("\U003C", "0.001", sep = "")
    else if (x < 0.01)
        sprintf("=%.3f", x)
    else
        sprintf("=%.2f", x)
}
```

```{r}
LBBB <- filter(dillacs, lbbb1_control0 == "LBBB")
Control <- filter(dillacs, lbbb1_control0 == "Control")

print_m_iqr <- function(var, data) {
    d <- data[[var]]
    m <- median(d)
    q1 <- quantile(d, 0.25)
    q3 <- quantile(d, 0.75)
    
    if (var == "cure") {
        sprintf("%.2f [%.2f-%.2f]", m, q1, q3)
    } else {
        sprintf("%.1f [%.1f-%.1f]", m, q1, q3)
    }
}

print_auc_ci <- function(var) {
    auc <- stats[[var]]$auc
    ci_lower <- stats[[var]]$auc_ci["2.5%"]
    ci_upper <- stats[[var]]$auc_ci["97.5%"]
    sprintf("%.2f [%.2f-%.2f]", auc, ci_lower, ci_upper)
}

cure_wilcox_pval <- (wilcox.test(cure ~ lbbb1_control0, data = dillacs))$p.value
ssi_wilcox_pval  <- (wilcox.test(ssi ~ lbbb1_control0, data = dillacs))$p.value

as_percent <- function(x) 100*x

stats$cure$sens <- as_percent(stats$cure$sens)
stats$cure$spec <- as_percent(stats$cure$spec)
stats$cure$cisens <- as_percent(stats$cure$cisens)
stats$cure$cispec <- as_percent(stats$cure$cispec)
stats$ssi$sens <- as_percent(stats$ssi$sens)
stats$ssi$spec <- as_percent(stats$ssi$spec)
stats$ssi$cisens <- as_percent(stats$ssi$cisens)
stats$ssi$cispec <- as_percent(stats$ssi$cispec)

```

\newpage

# Abstract

*Background:* Cardiac imaging-based indices of left ventricular (LV) mechanical
dyssynchrony have limited accuracy for predicting the response to cardiac
resynchronization therapy (CRT). The aim of the study was to evaluate the
diagnostic performance of mechanical dyssynchrony indices in a study population
of patients with severely reduced ejection fraction and no LV myocardial scar
assessed by cardiovascular magnetic resonance (CMR), and either left bundle
branch block (LBBB) or normal QRS duration.
*Methods:* We retrospectively identified `r length(unique(dillacs$id))`
patients from three centers, with LV ejection fraction ${\le}{35}$%, no scar by
CMR late gadolinium enhancement, and either normal electrocardiographic QRS
duration (${<}120$ms) and normal frontal plane electrical axis (-30 to
+90 degrees) (control, `r sprintf("n=%d", nrow(Control))`), or LBBB by Strauss'
criteria (LBBB, `r sprintf("n=%d", nrow(LBBB))`). The CMR image data from these
subjects is made publicly available as part of this publication. CMR feature
tracking was used to derive circumferential strain in a midventricular
short-axis cine image. Using circumferential strain, mechanical dyssynchrony
was quantified as the circumferential uniformity ratio estimate (CURE) and the
systolic stretch index (SSI), respectively.
*Results:* Both CURE and SSI resulted in measures of mechanical dyssynchrony that were more
severe (lower CURE, higher SSI) in LBBB compared to controls
(CURE, median [interquartile range], `r print_m_iqr("cure", LBBB)` vs `r print_m_iqr("cure", Control)`,
p`r pval_print(cure_wilcox_pval)`; 
SSI `r print_m_iqr("ssi", LBBB)` vs `r print_m_iqr("ssi", Control)`,
p`r pval_print(ssi_wilcox_pval)`). 
SSI outperformed CURE in the ability to discriminate between LBBB and controls
(area under the receiver operating characteristics curve [95% confidence interval]
`r print_auc_ci("ssi")` vs `r print_auc_ci("cure")`, 
p`r pval_print(stats$compare_auc$p.value)`; 
sensitivity `r sprintf("%.0f [%.0f-%.0f]", stats$ssi$sens, stats$ssi$cisens["2.5%"], stats$ssi$cisens["97.5%"])` 
vs `r sprintf("%.0f [%.0f-%.0f] %%", stats$cure$sens, stats$cure$cisens["2.5%"], stats$cure$cisens["97.5%"])`,
`r sprintf("p%s", pval_print(stats$sens_xmcnemar))`; specificity `r sprintf("%.0f [%.0f-%.0f]", stats$ssi$spec, stats$ssi$cispec["2.5%"], stats$ssi$cispec["97.5%"])`
vs `r sprintf("%.0f [%.0f-%.0f] %%", stats$cure$spec, stats$cure$cispec["2.5%"], stats$cure$cispec["97.5%"])`,
`r sprintf("p%s", pval_print(stats$spec_xmcnemar))`).
*Conclusions:* The ability to discriminate between LBBB and normal QRS duration
among patients with severely reduced ejection fraction and no scar was fair for
CURE and excellent for SSI.