---
title: "Diagnostic performance of left ventricular mechanical dyssynchrony indices using cardiovascular magnetic resonance feature tracking"
author:
 - "Daniel E Loewenstein MD^1^, Björn Wieslander MD PhD^1^, Einar Heiberg PhD^2^, Jimmy Axelsson MD^1^, Igor Klem MD^3^, Robin Nijveldt MD PhD^4^, Erik B Schelbert MSc MD^5^, Peder Sörensson MD PhD^6^, Andreas Sigfridsson PhD^1^, David G Strauss MD^7^,\nRaymond J Kim MD^8^, Brett D Atwater MD^3,9^, Martin Ugander MD PhD^1,10^"
 - "^1^Department of Clinical Physiology, Karolinska University Hospital, and Karolinska Institutet, Stockholm, Sweden"
 - "^2^Clinical Physiology, Department of Clinical Sciences Lund, Lund University, Skåne University Hospital, Lund, Sweden"
 - "^3^Division of Cardiology, Duke University Medical Center, Durham, NC, USA"
 - "^4^Department of Cardiology, Radboud University Medical Center, Nijmegen, the Netherlands"
 - "^5^Minneapolis Heart Institute East, Allina Health, United Hospital, St. Paul, MN, USA"
 - "^6^Department of Medicine Solna, Karolinska Institutet, and Department of Cardiology, Karolinska University Hospital, Solna, Stockholm, Sweden"
 - "^7^Division of Applied Regulatory Science, Office of Clinical Pharmacology, Office of Translational Sciences, Center for Drug Evaluation and Research, U.S. Food and Drug Administration, Silver Spring, MD, USA"
 - "^8^Duke Cardiovascular Magnetic Resonance Center, Duke University Medical Center, Durham, NC, USA"
 - "^9^Section of Cardiac Electrophysiology, Inova Heart and Vascular Institute, Fairfax, VA, USA"
 - "^10^Kolling Institute, Royal North Shore Hospital, and University of Sydney, Sydney, Australia"
csl: auxilliary/vancouver.csl
bibliography: auxilliary/library-bibtex-version.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```

```{r}
# Set paths
PROJECT_DIR <- rprojroot::find_rstudio_root_file()
RAW_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "raw_data")
DERIVED_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "derived_data")
FIG_DIR <- file.path(PROJECT_DIR, "analysis", "figures")
```

```{r}
dillacs <- readRDS(file.path(DERIVED_DATA_DIR, "dyssynchrony_data.Rds"))
stats <- readRDS(file.path(DERIVED_DATA_DIR, "statistics.Rds"))
figs  <- readRDS(file.path(DERIVED_DATA_DIR, "figs.Rds"))
source(file.path(PROJECT_DIR, "R", "offset2groups.R"))
```

```{r}
pval_print <- function(x) {
    if (x < 0.001)
        paste("\U003C", "0.001", sep = "")
    else if (x < 0.01)
        sprintf("=%.3f", x)
    else
        sprintf("=%.2f", x)
}
```

```{r}
LBBB <- filter(dillacs, lbbb1_control0 == "LBBB")
Control <- filter(dillacs, lbbb1_control0 == "Control")

print_m_iqr <- function(var, data) {
    d <- data[[var]]
    m <- median(d)
    q1 <- quantile(d, 0.25)
    q3 <- quantile(d, 0.75)
    
    if (var == "cure") {
        sprintf("%.2f [%.2f-%.2f]", m, q1, q3)
    } else {
        sprintf("%.1f [%.1f-%.1f]", m, q1, q3)
    }
}

print_auc_ci <- function(var) {
    auc <- stats[[var]]$auc
    ci_lower <- stats[[var]]$auc_ci["2.5%"]
    ci_upper <- stats[[var]]$auc_ci["97.5%"]
    sprintf("%.2f [%.2f-%.2f]", auc, ci_lower, ci_upper)
}

cure_wilcox_pval <- (wilcox.test(cure ~ lbbb1_control0, data = dillacs))$p.value
ssi_wilcox_pval  <- (wilcox.test(ssi ~ lbbb1_control0, data = dillacs))$p.value

as_percent <- function(x) 100*x

stats$cure$sens <- as_percent(stats$cure$sens)
stats$cure$spec <- as_percent(stats$cure$spec)
stats$cure$cisens <- as_percent(stats$cure$cisens)
stats$cure$cispec <- as_percent(stats$cure$cispec)
stats$ssi$sens <- as_percent(stats$ssi$sens)
stats$ssi$spec <- as_percent(stats$ssi$spec)
stats$ssi$cisens <- as_percent(stats$ssi$cisens)
stats$ssi$cispec <- as_percent(stats$ssi$cispec)

```

\newpage

# Abstract

A subgroup of patients with non-specific intraventricular conduction delay
(IVCD) have a left bundle branch block (LBBB)-like activation pattern that could
potentially benefit from cardiac resynchronization therapy. However, wihle
indices of dyssynchrony in LBBB could potentially be applied to other patient
groups, their relative accuracies in LBBB are not known. Therefore, we evaluated
the classification accuracy of dyssynchrony indices in a study population of
patients with severely reduced ejection fraction and no scar by cardiovascular
magnetic resonance (CMR), and either left bundle branch block (LBBB,
`r sprintf("n=%d", nrow(LBBB))`) or normal QRS duration (control, `r sprintf("n=%d", nrow(Control))`).
This CMR image data is made publicly
available. Using CMR feature tracking circumferential strain, dyssynchrony was
quantified as the circumferential uniformity ratio estimate (CURE) and the
systolic stretch index (SSI). Both CURE and SSI quantified more severe
dyssynchrony in LBBB compared to controls (p${<}0.001$ for booth). SSI
outperformed CURE for discriminating between LBBB and controls (area under the
receiver operating characteristics curve [95% confidence interval]
`r print_auc_ci("ssi")` vs `r print_auc_ci("cure")`, 
p`r pval_print(stats$compare_auc$p.value)`). The ability to
discriminate between LBBB and normal QRS duration among patients with severely
reduced ejection fraction and no scar was fair for CURE and excellent for SSI.
SSI merits further evaluation for detecting dyssynchrony in IVCD.

# Keywords

Cardiac Magnetic Resonance; Mechanical dyssynchrony; Left Bundle Branch Block;
Cardiac Resynchronization Therapy; Heart Failure; Circumferential Strain;
Feature Tracking