---
title: "Accuracy of left ventricular mechanical dyssynchrony indices for mechanical characteristics of left bundle branch block using cardiovascular magnetic resonance feature tracking"
author:
 - "Daniel E Loewenstein MD^1^, Björn Wieslander MD PhD^1^, Einar Heiberg PhD^2^, Jimmy Axelsson MD^1^, Igor Klem MD^3^, Robin Nijveldt MD PhD^4^, Erik B Schelbert MSc MD^5^, Peder Sörensson MD PhD^6^, Andreas Sigfridsson PhD^1^, David G Strauss MD^7^,\nRaymond J Kim MD^8^, Brett D Atwater MD^9^, Martin Ugander MD PhD^1,10^"
 - "^1^Department of Clinical Physiology, Karolinska University Hospital, and Karolinska Institutet, Stockholm, Sweden"
 - "^2^Clinical Physiology, Department of Clinical Sciences Lund, Lund University, Skåne University Hospital, Lund, Sweden"
 - "^3^Division of Cardiology, Duke University Medical Center, Durham, NC, USA"
 - "^4^Department of Cardiology, Radboud University Medical Center, Nijmegen, the Netherlands"
 - "^5^Minneapolis Heart Institute East, Allina Health, United Hospital, St. Paul, MN, USA"
 - "^6^Department of Medicine Solna, Karolinska Institutet, and Department of Cardiology, Karolinska University Hospital, Solna, Stockholm, Sweden"
 - "^7^Division of Applied Regulatory Science, Office of Clinical Pharmacology, Office of Translational Sciences, Center for Drug Evaluation and Research, U.S. Food and Drug Administration, Silver Spring, MD, USA"
 - "^8^Duke Cardiovascular Magnetic Resonance Center, Duke University Medical Center, Durham, NC, USA"
 - "^9^Section of Cardiac Electrophysiology, Inova Schar Heart and Vascular, Fairfax, VA, USA"
 - "^10^Kolling Institute, Royal North Shore Hospital, and University of Sydney, Sydney, Australia"
csl: auxilliary/vancouver.csl
bibliography: auxilliary/library-bibtex-version.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```

```{r}
# Set paths
PROJECT_DIR <- rprojroot::find_rstudio_root_file()
RAW_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "raw_data")
DERIVED_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "derived_data")
FIG_DIR <- file.path(PROJECT_DIR, "analysis", "figures")
```

```{r}
dillacs <- readRDS(file.path(DERIVED_DATA_DIR, "dyssynchrony_data.Rds"))
stats <- readRDS(file.path(DERIVED_DATA_DIR, "statistics.Rds"))
figs  <- readRDS(file.path(DERIVED_DATA_DIR, "figs.Rds"))
source(file.path(PROJECT_DIR, "R", "offset2groups.R"))
```

```{r}
pval_print <- function(x) {
    if (x < 0.001)
        paste("\U003C", "0.001", sep = "")
    else if (x < 0.01)
        sprintf("=%.3f", x)
    else
        sprintf("=%.2f", x)
}
```

```{r}
LBBB <- filter(dillacs, lbbb1_control0 == "LBBB")
Control <- filter(dillacs, lbbb1_control0 == "Control")

print_m_iqr <- function(var, data) {
    d <- data[[var]]
    m <- median(d)
    q1 <- quantile(d, 0.25)
    q3 <- quantile(d, 0.75)
    
    if (var == "cure") {
        sprintf("%.2f [%.2f-%.2f]", m, q1, q3)
    } else {
        sprintf("%.1f [%.1f-%.1f]", m, q1, q3)
    }
}

print_auc_ci <- function(var) {
    auc <- stats[[var]]$auc
    ci_lower <- stats[[var]]$auc_ci["2.5%"]
    ci_upper <- stats[[var]]$auc_ci["97.5%"]
    sprintf("%.2f [%.2f-%.2f]", auc, ci_lower, ci_upper)
}

cure_wilcox_pval <- (wilcox.test(cure ~ lbbb1_control0, data = dillacs))$p.value
ssi_wilcox_pval  <- (wilcox.test(ssi ~ lbbb1_control0, data = dillacs))$p.value

as_percent <- function(x) 100*x

stats$cure$sens <- as_percent(stats$cure$sens)
stats$cure$spec <- as_percent(stats$cure$spec)
stats$cure$cisens <- as_percent(stats$cure$cisens)
stats$cure$cispec <- as_percent(stats$cure$cispec)
stats$ssi$sens <- as_percent(stats$ssi$sens)
stats$ssi$spec <- as_percent(stats$ssi$spec)
stats$ssi$cisens <- as_percent(stats$ssi$cisens)
stats$ssi$cispec <- as_percent(stats$ssi$cispec)

```

\newpage

# Abstract
**Aims:** More than 90% of patients with left bundle branch block (LBBB) and reduced left
ventricular (LV) ejection fraction have LV dyssynchrony and a high probability
of response to cardiac resynchronization therapy (CRT). A subgroup of patients
with non-specific intraventricular conduction delay (IVCD) have a LBBB-like LV
activation pattern when studied using invasive mapping and advanced
echocardiographic techniques. These patients also frequently benefit from CRT
but these patients have proven difficult to identify using ECG criteria.
Cardiovascular magnetic resonance (CMR) imaging indices of dyssynchrony may
identify patients with IVCD who may benefit from CRT but their relative
accuracies for identification of LV dyssynchrony remains unknown. We compared the
LV dyssynchrony classification accuracy of two commonly available CMR indices
in a study population of patients with severely reduced LV ejection fraction
and no scar, and either LBBB or normal conduction (normal QRS duration and
axis, controls).

**Methods and results:** In LBBB (`r sprintf("n=%d", nrow(LBBB))`) and controls
(`r sprintf("n=%d", nrow(Control))`), using CMR feature tracking circumferential
strain, dyssynchrony was quantified as the circumferential uniformity ratio
estimate (CURE) and the systolic stretch index (SSI). Deidentified CMR
image-data were made publicly available. Both CURE and SSI quantified more
severe dyssynchrony in LBBB compared to controls (p${<}0.001$ for both). SSI
more frequently discriminated LBBB and normal conduction LV activation patterns
than CURE (area under the receiver operating characteristics curve [95%
confidence interval] `r print_auc_ci("ssi")` for SSI vs `r print_auc_ci("cure")`
for CURE, p`r pval_print(stats$compare_auc$p.value)`).

**Conclusion:** SSI is superior to CURE for discriminating synchronous and
dyssynchronous LV activation and should be further studied in the setting of
non-LBBB conduction abnormalities.

**Keywords**: Cardiac Magnetic Resonance; Mechanical dyssynchrony; Left Bundle
Branch Block; Cardiac Resynchronization Therapy; Heart Failure; Circumferential
Strain; Feature Tracking