```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```


```{r}
# Set paths
PROJECT_DIR <- rprojroot::find_rstudio_root_file()
RAW_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "raw_data")
DERIVED_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "derived_data")
FIG_DIR <- file.path(PROJECT_DIR, "analysis", "figures")
```

```{r}
dillacs <- readRDS(file.path(DERIVED_DATA_DIR, "dyssynchrony_data.Rds"))
stats <- readRDS(file.path(DERIVED_DATA_DIR, "statistics.Rds"))
figs  <- readRDS(file.path(DERIVED_DATA_DIR, "figs.Rds"))
source(file.path(PROJECT_DIR, "R", "offset2groups.R"))
```

```{r}
pval_print <- function(x) {
    if (x < 0.001)
        paste("\U003C", "0.001", sep = "")
    else if (x < 0.01)
        sprintf("=%.3f", x)
    else
        sprintf("=%.2f", x)
}
```

```{r}
# Taken from LP-figures

format_thresh <- function(x) {
    if (x < 1)
        sprintf("Threshold: %.2f", x)
    else
        sprintf("Threshold: %.1f", x)
}

format_thresh <- Vectorize(format_thresh)

text_height <- 0.25
text_size   <- 4
```

```{r}
LBBB <- filter(dillacs, lbbb1_control0 == "LBBB")
Control <- filter(dillacs, lbbb1_control0 == "Control")

as_percent <- function(x) 100*x

stats$cure$sens <- as_percent(stats$cure$sens)
stats$cure$spec <- as_percent(stats$cure$spec)
stats$cure$cisens <- as_percent(stats$cure$cisens)
stats$cure$cispec <- as_percent(stats$cure$cispec)
stats$ssi$sens <- as_percent(stats$ssi$sens)
stats$ssi$spec <- as_percent(stats$ssi$spec)
stats$ssi$cisens <- as_percent(stats$ssi$cisens)
stats$ssi$cispec <- as_percent(stats$ssi$cispec)

get_frequent <- function(x) {
    if (is.character(x))
        res <- table(unlist(dillacs[[x]]))
    else
        res <- table(unlist(x))
    
    as.numeric(names(res[which.max(res)]))
}

img_prms <- vector("list", 6)
names(img_prms) <- c("TE", "TR", "flip_angle", "slice_mm", "t_res", "segment_version")

img_prms$TE <- get_frequent("echo_time")
img_prms$TR <- get_frequent("rep_time")
img_prms$flip_angle <- get_frequent("flip_angle")
img_prms$slice_mm   <- get_frequent("slice_thickness")
img_prms$t_res      <- get_frequent(dillacs$t_max / unlist(dillacs$t_incr))
img_prms$segment_version <- dillacs$program_version[[1]]
```

# Methods

This is an observational case-control study where patients were
retrospectively identified by cross-referencing the CMR and electrocardiography
(ECG) databases from three centers (Duke University Medical Center, NC, USA;
Pittsburgh University Medical Center, PA, USA; and Karolinska University
Hospital, Stockholm, Sweden). The study was approved by the local
human subject research ethics committee at each site, and all subjects either
provided written informed consent or were included following a retrospective
waiver of informed consent provided by the local ethics committee.

## Subject selection

Subjects considered for inclusion in the present study had a LV
ejection fraction ${\le}35\%$, no scar by CMR late gadolinium enhancement
(LGE), CMR cine images in a LV short-axis stack, and either normal ECG QRS
duration (${<}120$ ms) and frontal plane electrical axis (-30 to +90 degrees,
controls, n=`r nrow(Control)`), or LBBB (n=`r nrow(LBBB)`) defined by Strauss'
strict ECG criteria, defined as a terminal negative deflection in lead $V_1$
and $V_2$ (QS or rS configuration), a QRS duration ${\ge}140$ ms for men and
${\ge}130$ ms for women, and the presence of mid-QRS notching or slurring in 
${\ge}2$ of leads $V_1,~V_2,~V_5,~V_6$, I and aVL [@strauss_definingleftbundle_2011].
Subjects were excluded if they had a history of congenital heart disease, CMR evidence
of myocardial storage disease, atrial fibrillation, prior open heart surgery,
or LV septal wall flattening indicative of clinically significant pulmonary
hypertension. The following baseline characteristics were collected: age, sex,
height, weight, body surface area (BSA), body mass index (BMI), and CMR
measures of LV volumes, function and mass. Among the patients who met the
inclusion criteria (n=87), patients were excluded due to having Takotsubo
cardiomyopathy (n=1), atrial fibrillation discovered at time of feature
tracking analysis (n=1), or missing or insufficient number of diagnostic quality
CMR cine images (n=5). As a result, the final study group included
`r nrow(dillacs)` patients.

## CMR image acquisition

All imaging was performed with clinically available scanners at the respective
centers. Scanners included 3T (Siemens Verio, Erlangen, Germany) and 1.5T
systems (Siemens Avanto, Espree or Aera, Erlangen, Germany, or Philips Intera,
Best, the Netherlands), all using ECG gating and phased-array receiver coils.
Typical acquisition parameters for cine images were: repetition time `r round(img_prms$TR,0)` ms,
echo time `r round(img_prms$TE, 1)` ms, flip angle `r img_prms$flip_angle` degrees, 
matrix 190 x 190, slice thickness `r img_prms$slice_mm` mm,
and temporal resolution `r img_prms$t_res` frames per cardiac cycle. 
Clinical reports of cardiac viability assessment were reviewed for mention of 
any myocardial scar by LGE.

## Image and strain analysis

Cine CMR images exported for offline myocardial strain analysis performed by an
observer using commercially available software for CMR feature
tracking (Segment version `r img_prms$segment_version`) Medviso, Lund, Sweden)
[@morais_cardiovascularmagneticresonance_2017; @heyde_elastic_2013]. All analysis was performed
blinded to ECG classification. Endocardial and epicardial borders, excluding
papillary muscles and trabeculations, were manually delineated in the
end-diastolic reference timeframe. The end-diastolic reference timeframe was
set to the timeframe immediately following halting of the circumferential
expansion and longitudinal lengthening of the LV during late diastole as
viewed in a three chamber, long-axis, and short-axis slice. This was due to
observation that there was often a delay in the closure of the mitral valve,
which otherwise has commonly been used to define end-diastole. The delineation was
performed in a single midventricular short-axis slice. A non-rigid elastic
registration strategy was used by the software to measure myocardial strain over
time. For regional strain assessment, the area encompassed by the endo- and
epicardial borders was segmented into regions of interest according six segments
of the American Heart Association 17-segment model. In short-axis images, the
location of regional segments was determined using an angle relative to the
right ventricular anterior insertion point. Circumferential strain was
evaluated from the Lagrangian strain tensor between adjacent points. Mechanical
dyssynchrony was quantified using the circumferential uniformity ratio estimate
(CURE) [@helm_cardiac_2005; @leclercq_systolicimprovementmechanical_2002;
@bilchick_cardiac_2008]
and the systolic stretch index (SSI)
[@gorcsan_systolicstretchcharacterizes_2018; @lumens_differentiatingelectromechanicalnon_2015].
In short, CURE is derived
from Fourier transformation of the spatial distribution of strain from
myocardial segments averaged over the number of short axis slices. CURE is then
calculated as
$$CURE = \frac{1}{n} \sum_{t=1}^{n}{\sqrt{\frac{\sum{S_0(t)}}{\sum{S_0(t)} + \sum{S_1(t)}}}}$$
where $S_0$ is the zero order, and $S_1$ is the first order term in the fourier
transformation, and $n$ is the number of timeframes covering the cardiac cycle.
CURE ranges between 0 (perfect dyssynchrony) and 1 (perfect synchrony).
SSI was originally developed through computer simulations
[@lumens_differentiatingelectromechanicalnon_2015], and later presented in a
slightly simplified version for use in echocardiography [@gorcsan_systolicstretchcharacterizes_2018] and is 
calculated as the sum of LV lateral wall systolic pre-stretch
(SPS) and septal rebound stretch (SRS).
$$SSI = \frac{SPS_{antlat} + SPS_{inflat}}{2} + \frac{SRS_{antsept} + SRS_{infsept}}{2}$$
Systolic pre-stretch is defined as the sum of LV lateral wall stretch before aortic
valve opening, averaged over the anterolateral and inferolateral segment. Septal
rebound stretch is defined as the sum of septal stretch following early systolic
shortening and before aortic valve closure, averaged over the anteroseptal and 
inferoseptal segment.

## Statistical analysis

Categorical data are reported as number and percentages. Continuous variables
are reported as median [interquartile range]. For continuous variables, groups
were compared using the Wilcoxon signed-rank test. Bivariate correlation was
examined by Spearman's $\rho$ correlation coefficient.
Univariable logistic regression models with LBBB status as the dependent
variable were fitted separately for the two dyssynchrony indices. Specificity,
sensitivity, discriminatory performance, and cut-off values were derived from
receiver-operating characteristics (ROC) analysis using the Youden's index.
Multivariable linear regression models fit separately for each dyssynchrony
index in LBBB, and controls respectively, were used to test for associations
with covariates age, LV end-diastolic volume index (LVEDVI), LV mass index
(LVMI), and sex, indicating any need for covariate-adjusted, or
covariate-specific ROC curves. Nonlinearities were entertained by use of
restricted cubic splines, as were interactions between LVEDVI, and LVMI with
sex, respectively. Areas under the paired ROC curves were compared using
nonparametric stratified bootstrapping. Bootstrapped CIs were derived from 4000
replicates and calculated using the percentile method. A two-sided $p < 0.05$
was considered statistically significant. Data processing and statistical
analysis were performed in the R statistical programming environment
`r getRversion()` [@rcoreteam_r:_2017], using package dplyr 
`r packageVersion("dplyr")` [@wickham_dplyr:_2017] for data transformation, ggplot2 
`r packageVersion("ggplot2")` [@wickham_ggplot2:_2009] for graphical
visualizations, pROC `r packageVersion("pROC")` [@robin_proc:_2011] for ROC
analysis, rms `r packageVersion("rms")` for regression modeling, and knitr 
`r packageVersion("knitr")` [@xie_knitr:_2017] for reproducible documentation.
