```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(Hmisc)
library(rms)
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r}
# Set paths
PROJECT_DIR <- rprojroot::find_rstudio_root_file()
RAW_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "raw_data")
DERIVED_DATA_DIR <- file.path(PROJECT_DIR, "analysis", "data", "derived_data")
FIG_DIR <- file.path(PROJECT_DIR, "analysis", "figures")
```

```{r}
dillacs <- readRDS(file.path(DERIVED_DATA_DIR, "dyssynchrony_data.Rds"))
dillacs <- dillacs %>% 
    mutate(lbbb1_control0 = factor(lbbb1_control0, levels = c("Control", "LBBB")))

stats <- readRDS(file.path(DERIVED_DATA_DIR, "statistics.Rds"))
figs  <- readRDS(file.path(DERIVED_DATA_DIR, "figs.Rds"))
source(file.path(PROJECT_DIR, "R", "offset2groups.R"))
```

```{r}
format_thresh <- function(x, xmin, xmax) {
    if (x < 1)
        sprintf("Threshold: %.2f [%.2f, %.2f]", x, xmin, xmax)
    else
        sprintf("Threshold: %.1f [%.1f, %.1f]", x, xmin, xmax)
}

format_thresh <- Vectorize(format_thresh)

text_height <- 0.25
text_size   <- 4
safe_colors <- RColorBrewer::brewer.pal(3, "PuOr")[c(1, 3)]
```

# Results

```{r}
male_n <- c(sprintf("%s (%s%%)", sum(dillacs$Gender=="M"), (sum(dillacs$Gender == "M")/length(dillacs$Gender))*100))

female_perc <- c(sprintf("%.0f%%", (sum(dillacs$Gender ==
"F")/length(dillacs$Gender))*100))

sumstats <- summaryM(Age + cure + ssi + EDVI + LVMI + Gender + BMI + EDV + LVEF + QRSDur
              + LVM ~ lbbb1_control0, 
              data = dillacs, test = TRUE, overall = TRUE)

get_pval <- function(x, data = sumstats) {
    format.pval(data$results$.ALL.$testresults[[c(deparse(substitute(x)), "P")]], eps=0.001)
}

get_median <- function(x, group, dig = 1, data = sumstats) {
    round(data$results$.ALL.$stats[[deparse(substitute(x))]][group, "0.500"], dig)
}

print_iqr <- function(x, group=3, dig = 1, data = sumstats) {
    sprintf(paste0("[%.", dig,"f-%.",dig,"f]"),
            data$results$.ALL.$stats[[deparse(substitute(x))]][group, "0.250"],
            data$results$.ALL.$stats[[deparse(substitute(x))]][group, "0.750"])
}

```

## Subject characteristics

The characteristics of patients included in the study 
(n=`r nrow(dillacs)`, `r female_perc` female) are presented in
Table \@ref(tab:tableone). Characteristics were similar in the two groups except for
older age and greater LV mass in subjects with LBBB.

```{r}
dysstats <- summaryM(cure + ssi ~ lbbb1_control0, 
              data = dillacs, test = TRUE, overall = TRUE)

formals(get_pval) <- alist(x=, data = dysstats)
formals(get_median) <- alist(x=, group=, dig = 1, data = dysstats)
formals(print_iqr) <- alist(x=, group = 3, dig = 1, data = dysstats)

get_OR <- function(coef, data, inv = FALSE, dig = 2) {
    res <- exp(data[["coef.ci"]][deparse(substitute(coef)), "coef"])
    if(inv) {
        res <- 1/res
    }
   format(round(res, dig), nsmall = 2)
}

get_OR_ci <- function(coef, data, inv = FALSE, dig = 2) {
    res <- exp(data[["coef.ci"]][deparse(substitute(coef)), c("2.5%", "97.5%")])
    if(inv) {
        res <- 1/res
        res <- c(res[2], res[1])
    }
    sprintf(paste0("[95%% CI: %.", dig,"f-%.",dig,"f", "]"), res[1], res[2])
}

print_cor <- function(group, var) {
    group = deparse(substitute(group))
    var   = deparse(substitute(var))
    r <- stats[[var]]$cor[[group]]$estimate
    ci <- stats[[var]]$cor[[group]]$conf.int
    p  <- stats[[var]]$cor[[group]]$p.value
    p   <- ifelse(p < 0.001, 
                  format.pval(p, eps = 0.001, digits = 3),
                  paste0("=", format.pval(p, eps = 0.001, digits = 3))
                  )

    sprintf("$r_{s}$ = %.2f [95%% CI: %.2f-%.2f]; p%s",
            r, ci[1], ci[2], p)
}

print_median_ci <- function(var) {
    if(var == "cure") {
        sprintf("%.2f [95%% CI: %.2f-%.2f]",
                stats[[var]]$dys_median_ci$t0,
                stats[[var]]$dys_median_ci$percent[4],
                stats[[var]]$dys_median_ci$percent[5])
    } else {
        sprintf("%.1f [95%% CI: %.1f-%.1f]",
                stats[[var]]$dys_median_ci$t0,
                stats[[var]]$dys_median_ci$percent[4],
                stats[[var]]$dys_median_ci$percent[5])
    }
}

print_lrt <- function(var) {
    chi <- stats[[var]]$modstats["Model L.R."]
    df  <- stats[[var]]$modstats["d.f."]
    p   <- stats[[var]]$modstats["P"]
    p   <- ifelse(p < 0.001, 
                  format.pval(p, eps = 0.001, digits = 3),
                  paste0("=", format.pval(p, eps = 0.001, digits = 3))
                  )
    sprintf("$\\chi^2(%.1f)$, $df$=%.0f, p%s", chi, df, p)
}

print_mchisq <- function(var) {
    chi <- stats[[var]]$fit.anova[" Nonlinear", "Chi-Square"]
    df  <- stats[[var]]$fit.anova[" Nonlinear","d.f."]
    p   <- stats[[var]]$fit.anova[" Nonlinear","P"]
    p   <- ifelse(p < 0.001,
                  format.pval(p, eps = 0.001, digits = 3),
                  paste0("=", format.pval(p, eps = 0.001, digits = 3))
                  )
    sprintf("$\\chi^2(%.1f)$, $df$=%.0f, p%s", chi, df, p)
}

print_chisq <- function(var) {
    res <- stats[[var]]$fit.anova[var, c("Chi-Square", "P")]
    p   <- ifelse(res[2] < 0.001,
                  format.pval(res[2], eps = 0.001, digits = 3),
                  paste0("=", format.pval(res[2], eps = 0.001, digits = 3))
                  )
    sprintf("%.2f (p%s)", res[1], p)
}

print_gof <- function(var) {
    ss <- stats[[var]]$fit.gof["Sum of squared errors"]
    p  <- stats[[var]]$fit.gof["P"]
    p  <- ifelse(p < 0.001,
                 format.pval(p, eps = 0.001, digits = 3),
                 paste0("=", format.pval(p, eps = 0.001, digits = 3))
                 )
    sprintf("%.2f (p%s)",ss, p)
}
```

```{r}
cure_loc_shift <- stats$cure$wilcox
ssi_loc_shift <- stats$ssi$wilcox
```


## Dyssynchrony measurements

Both CURE and SSI showed group differences between LBBB and controls (Figure \@ref(fig:dysboxplot)).
Consistent with a greater amount of mechanical dyssynchrony in LBBB, CURE was
lower in LBBB compared to controls
(`r get_median(cure, 2, 2)` `r print_iqr(cure, 2, 2)` vs  `r get_median(cure, 1, 2)` `r print_iqr(cure, 1, 2)`, `r sprintf("p<%.3f", ifelse(cure_loc_shift$p.value < 0.001, 0.001,  cure_loc_shift$p.value))`), 
and SSI was higher in LBBB compared to controls
(`r get_median(ssi, 2)` `r print_iqr(ssi, 2)` vs  `r get_median(ssi, 1)`
`r print_iqr(ssi, 1)`, 
`r sprintf("p<%.3f", ifelse(ssi_loc_shift$p.value < 0.001, 0.001, ssi_loc_shift$p.value))`).
Compared to CURE, SSI had a greater area under the ROC curve for detecting
mechanical dyssynchrony associated with strict LBBB 
(`r sprintf("%.2f [95%% CI: %.2f-%.2f]", stats$ssi$auc, stats$ssi$auc_ci["2.5%"], stats$ssi$auc_ci["97.5%"])`
vs
`r sprintf("%.2f [95%% CI: %.2f-%.2f]", stats$cure$auc, stats$cure$auc_ci["2.5%"], stats$cure$auc_ci["97.5%"])`,
p`r format.pval(stats$compare_auc$p.value, eps = 0.001)`, Figure \@ref(fig:dysrocplot)), and this corresponded to a higher sensitivity and specificity for SSI compared to CURE
(Figure \@ref(fig:performanceplot)). The odds ratio (OR) for identifying LBBB for SSI was `r get_OR(ssi, stats$ssi)`
`r get_OR_ci(ssi, stats$ssi)` per 1 percentage unit increase in SSI value, and for CURE was 
`r get_OR(cure, stats$cure)` `r get_OR_ci(cure, stats$cure)` per 
0.10 decrease in CURE value. In evaluating the need for covariate-adjusted and/or
covariate-specific ROC curves, linear regression models were used to test the
association between dyssynchrony indices and covariates: age, LVEDVI, LVMI, and
sex, allowing for interactions between LVMI and LVEDVI, with sex, respectively.
No evidence was found in support of an association between dyssynchrony indices
and covariates in controls, or between CURE and covariates in LBBB. Furthermore,
neither CURE nor SSI associated with QRS duration in either LBBB or controls
(p ${\ge}0.4$ for all). There was an association between SSI and age, keeping other
covariates fixed, in LBBB. This suggests that the discriminatory ability of SSI
might vary with respect to age. For purposes of clarity, the unadjusted ROC
curve is presented.

## CMR image availability

All CMR images analyzed as part of the current study are made available online
(doi: [10.6084/m9.figshare.15155596](https://doi.org/10.6084/m9.figshare.15155596)).
Among the individual CMR exams (n=80) the majority were performed on a Siemens
scanner (n=75) and only a few on a Philips scanner (n=5), all contain a cine
short-axis stack, while most also contain cine 2-chamber (n=78), 3-chamber
(n=79), and 4-chamber images (n=79). As per the inclusion criteria all patients
were verified to be free of myocardial scar by LGE CMR.
